---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# paleoxrf

The goal of paleoxrf is to simplify reading data from Acadia's paleoenvironmental X-Ray Fluorescence laboratory.

## Installation

You can install **paleoxrf** from GitHub using the following code:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("paleolimbot/paleoxrf")
```

If all is well, you should be able to load the package:

```{r}
library(paleoxrf)
```

## Reading XRF Data

The package provides a few functions to read XRF data output from several types of machines, such as the [Olympus Vanta](http://www.olympus-ims.com/en/vanta/). The function reads the raw output of the machines, rather than a "cleaned up" version from other spreadsheet software.

```{r}
test_file_oly <- system.file(
  "xrf_files/olympus_vanta_test.csv", 
  package = "paleoxrf"
)

read_olympus_vanta(test_file_oly)
```

```{r}
test_file_pan <- system.file(
  "xrf_files/panalytical_test.txt", 
  package = "paleoxrf"
)

read_panalytical_txt(test_file_pan)
```


