---
title: "Wrangling Paleo Data"
author: "Dewey Dunnington"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wrangling Paleo Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(tibble)
library(dplyr)
library(paleoxrf)
library(tidyr)
library(ggplot2)
library(purrr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5
)
```

This tutorial will be most useful with a basic understanding of data manipulation using the tidyverse. The tutorials from [R for Data Science](http://r4ds.had.co.nz/) on [data transformation](http://r4ds.had.co.nz/transform.html), [tidy data](http://r4ds.had.co.nz/tidy-data.html), and [data visualization](http://r4ds.had.co.nz/data-visualisation.html) are useful starting places.

First, load the packages:

```{r, eval=FALSE}
library(tidyverse)
library(paleoxrf)
```

## Working with XRF Data

Most XRF data frames are scarily huge, but thanks to the functions in the tidyverse, working with huge data frames is just as easy as working with small ones. For an example, we will use a subset of core data from Pockwock Lake and Lake Major. You can load this data into your workspace by typing `data(pocmaj_raw)`.

```{r}
data(pocmaj_raw)
head(pocmaj_raw)
```

```{r, eval = FALSE, echo = FALSE}
# above data in tribble form
pocmaj_raw <- tribble(
  ~sample_id, ~Ca, ~Ti, ~V,  
  "poc15-2 0",  1036, 1337, 29,
  "poc15-2 0", 1951, 2427, 31,
  "poc15-2 0", 1879, 2350, 39,
  "poc15-2 1", 1488, 2016, 36,
  "poc15-2 2", 2416, 3270, 79,
  "poc15-2 3", 2253, 3197, 79,
  "poc15-2 4", 2372, 3536, 87,
  "poc15-2 5", 2621, 3850, 86,
  "poc15-2 5", 2785, 3939, 95,
  "poc15-2 5", 2500, 3881, 80,
  "maj15-1 0", 1623, 2104, 73,
  "maj15-1 0", 1624, 2174, 73,
  "maj15-1 0", 2407, 2831, 89,
  "maj15-1 1", 1418, 2409, 70,
  "maj15-1 2", 1550, 2376, 70,
  "maj15-1 3", 1448, 2485, 64,
  "maj15-1 4", 1247, 2414, 57,
  "maj15-1 5", 1463, 1869, 78,
  "maj15-1 5", 1269, 1834, 71,
  "maj15-1 5", 1505, 1989, 94
)
```

### Wrangling the sample ID column

The first challenge is converting the `sample_id` column into a `core` and `depth` column, which are the two columns needed to properly plot the data. For this, we will use the `separate()` function within the tidyverse (for more advanced manipulation, see the `extract()` function).

```{r}
pocmaj_clean <- pocmaj_raw %>%
  separate(sample_id, into = c("core", "depth"), sep = " ")

head(pocmaj_clean)
```

The `separate()` function takes a data.frame and three arguments: the column containing the values to separate, the names of the output columns, and the separator to use. This is technically a [regular expression](https://en.wikipedia.org/wiki/Regular_expression), which will only matter if you need to split on a string that contains special characters such as `+\[]()?*.{}`. Usually this isn't a problem, but if it is you can escape the string with a backslash like this: `sep = "\\+"`. You can also keep your original `sample_id` column by passing `remove = FALSE`.

There is a good chance that some of your `sample_id` values will be misspelled for some reason or another. A simple way to fix these values is using `if_else()`, which can be used to replace specific values in a column.

```{r}
pocmaj_raw %>%
  mutate(sample_id = if_else(sample_id == "poc15-2 1", "the correct value", sample_id)) %>%
  head()
```

For more advanced manipulation, use the `stringr` package, which provides the function `str_replace()` (among others) that can perform search and replace queries along the column.

The next step is to convert depth values into numbers (they are currently text!). For this we will use `mutate()` and `as.numeric()`:

```{r}
pocmaj_clean <- pocmaj_raw %>%
  separate(sample_id, into = c("core", "depth"), sep = " ") %>%
  mutate(depth = as.numeric(depth))

head(pocmaj_clean)
```

### Parameter-long data

Occasionally, data in parameter-wide form (like the above) is useful, but to summarise replicates for a whole bunch of parameters and plot all the parameters at once, we need the data in parameter-long form. This form is more difficult to understand, but easier to work with! To convert the data to parameter-long form, we can use the `gather()` function.

```{r}
pocmaj_long <- pocmaj_clean %>%
  gather(Ca:V, key = "param", value = "value")

head(pocmaj_long)
```

The `gather()` function takes a data frame plus three arguments: the columns to gather, the `key` column (in which the column names are placed), and the `value` column (in which the values corresponding to each row/key combination are drawn). The columns not mentioned act as identifying variables that identify unique rows, which means that columns that contain measured values will cause problems! These can be removed using something like `select(-ends_with("_error"))`, or something similar. If you don't quite understand this step, bear with me, because it makes plotting and summarising a whole lot easier!

### Summarising replicates

The final step before plotting is to summarise replicate values. For this, we will use `group_by()` and `summarise()`.

```{r}
pocmaj_long_summarised <- pocmaj_long %>%
  group_by(core, depth, param) %>%
  summarise(mean_value = mean(value), sd_value = sd(value), n = n())

head(pocmaj_long_summarised)
```

Using `group_by()` then `summarise()` is common: `group_by()` specifies the columns whose unique combinations we are interested in. The values in these columns will identify unique rows in the output, which in our case are represented by `core`, `depth`, and `param`. The `summarise()` function takes arguments in the form of `output_column_name = expression`, where `expression` is an R expression (like `mean(value)`)) where column names can be used like variables. Using `mean()` and `sd()` is a good start, but `min()` and `max()` are also useful, as well as passing `na.rm = TRUE` if `NA` values exist in the `value` column.

### Converting parameter-long data to parameter-wide format

Some plotting functions and almost all ordination functions require data in parameter-wide. For this, we can use the opposite of `gather()`: the `spread()` function.

```{r}
pocmaj_wide_summarised <- pocmaj_long_summarised %>%
  select(core, depth, param, mean_value) %>%
  spread(key = param, value = mean_value)

head(pocmaj_wide_summarised)
```

## Plotting Paleo Data

There are several plotting libraries for paleo data, particularly for species composition data. These include [analogue](https://github.com/gavinsimpson/analogue) (the `Stratiplot()` function) and [rioja](https://cran.r-project.org/web/packages/rioja/index.html) (the `strat.plot()` function). For non-species data, the `ggplot2` library works quite well, provided data are in parameter-long form.

```{r}
ggplot(pocmaj_long_summarised, aes(y = depth, x = mean_value, colour = core)) +
  geom_errorbarh(
    aes(xmax = mean_value + sd_value, xmin = mean_value - sd_value),
    height = 0.1
  ) +
  geom_point() +
  geom_path() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse()
```

The `ggplot` library is quite intimidating at first, but it provides much flexibility and is worth the effort to [learn](http://r4ds.had.co.nz/data-visualisation.html). The above plot is constructed using a few lines, which I will describe one at a time.

```{r, eval = FALSE}
ggplot(pocmaj_long_summarised, aes(y = depth, x = mean_value, col = core)) +
```

The `ggplot()` function creates a plot using its first argument as the primary `data` source. In our case, this is the `pocmaj_long_summarised`. Within `ggplot()`, we specify the default **aesthetics**, which is a mapping between the columns in `data` and the information that `ggplot` needs to construct a plot. Generally, paleo diagrams have the depth on the y-axis, and the parameter value on the x-axis. If more than one value exists in `core` (or this column may not exist of the data is only for one core), we can use `colour = core` to plot each core using a different color.

```{r, eval = FALSE}
  geom_point() +
  geom_path() +
```

These `geom_*` functions don't need any arguments because they inherit the default aesthetics specified in `ggplot()`. We use `geom_path()` instead of `geom_line()` because `geom_line()` sorts its values by x value, which in our case doesn't make any sense!

```{r, eval = FALSE}
geom_errorbarh(
  aes(xmax = mean_value + sd_value, xmin = mean_value - sd_value),
  height = 0.1
) +
```

Including error information is important when constructing paleolimnological diagrams (when uncertainty information is available), which is why we include the fairly long call to `geom_errorbarh()`. Unlike `geom_point()` and `geom_path()`, error bars require more information than `x`, `y`, and `colour`. Instead, we need to specify additional aesthetics (`xmin` and `xmax`), and how these should be calculated given `data` (in our case, `xmax = mean_value + sd_value, xmin = mean_value - sd_value`). Finally, the `height` of the error bars needs to be adjusted or they look quite huge.

```{r, eval = FALSE}
facet_wrap(~param, scales = "free_x") +
```

The `facet_wrap()` specification is how `ggplot` creates many graphs using a single data input. The values in the specified column are used to create panels, with one panel per value. In our case, the `param` column is how we would like to separate our plots (this is usually the case). By default, `facet_wrap()` will keep all axes aligned, but because each parameter has a different range of values, we need that axis to automatically adjust based on its content. This is why we specify `scales = "free_x"`.

```{r, eval = FALSE}
  scale_y_reverse()
```

Finally, we need to reverse the y-axis for the traditional strat diagram look with zero depth at the top of the diagram. If ages are used, this can be omitted.

### Tips and tricks

There are as many variations on strat diagrams as there are cores in the world, and `ggplot` can't produce all of them. A few things are still possible with slight modification!

#### Dual Y axis (ages and depths)

Having a dual y axis is possible, but requires a function transforming depth to age. In this example I'll use a simple function assuming a constant sedimentation rate, but in reality this function will probably use the `approx()` function given known age/depth values from ^210^Pb or other dating method.

```{r}
depth_to_age <- function(depth, sed_rate = 0.1, core_year = 2017) {
  # sed_rate here is in cm/year
  core_year - 1 / sed_rate * depth
}
depth_to_age(0:10)
```

Given this function, it can be passed to the `trans` argument of `sec_axis` to create a secondary Y axis.

```{r}
ggplot(pocmaj_long_summarised, aes(y = depth, x = mean_value, col = core)) +
  geom_path() +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse(sec.axis = sec_axis(trans = ~depth_to_age(.), name = "age"))
```

The details of creating a secondary axis can be found in `?sec_axis`. Obviously this doesn't make sense for multiple cores, but works well for multiple parameters on a single core.

#### Zonation Lines

Zone lines can be added using `geom_hline()`, which creates a horizontal line superimposed over the plot.

```{r}
ggplot(pocmaj_long_summarised, aes(y = depth, x = mean_value, col = core)) +
  geom_path() +
  geom_hline(yintercept = c(1.8, 4.2), linetype = 2, alpha = 0.5) +
  facet_wrap(~param, scales = "free_x") +
  scale_y_reverse()
```

#### "Everything vs. everything"

Making a series of biplots is often useful, especially when dealing with XRF data. This is easy with data in its original, wide, format, but ggplot needs data in parameter-long form to make use of facets. This paired, long-form data can be created using a self `full_join()` using the non-parameter ID columns:

```{r}
long_pairs <- full_join(
  pocmaj_long_summarised,
  pocmaj_long_summarised,
  by = c("core", "depth")
)

head(long_pairs)
```

Creating a plot using this is quite straightforward (note that in this form, error bars can also be included using `geom_errorbar()` and `geom_errorbarh()`):

```{r}
ggplot(long_pairs, aes(x = mean_value.x, y = mean_value.y, col = core)) +
  geom_point() +
  facet_grid(param.y ~ param.x, scales = "free")
```

This data format has the added advantage of being able to test all the correlations for significance:

```{r}
long_pairs %>%
  group_by(param.x, param.y) %>%
  summarise(test = list(cor.test(mean_value.x, mean_value.y))) %>%
  mutate(test = map(test, broom::glance)) %>%
  unnest()
```

## Summary

The tidyverse offers a great number of possibilities with respect to core data, only a few of which I describe here. In general, the functions in the tidyverse allow for parameter-long data to be manipulated more easily, which allows for a greater amount of metadata and uncertainty information to be kept alongside the data until it is not needed. For more, see the extensive [tidyverse documentation](http://www.tidyverse.org/articles/) and companion book, [R for Data Science](http://r4ds.had.co.nz/).
